Speed and Precision in Range Analysis

### 2012-05-26 doc/Manuscript/SPE/paper.tex
Last updates in the paper.

We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
variable.
As we will show in Section~\ref{sub:micro}, once the growth pattern of $v$ is
known, we can replace $\fun{ft}(v)$ by an actual value.
After splitting the live ranges according to the rules stated above, we might
have to insert $\phi$-functions into the transformed program to re-convert it to
SSA form.
This last step avoids that two different names given to the same original
variable be simultaneously alive at the program code.
Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
standard e-SSA form.
We have not created variable names for $i_f$ and $j_f$, because neither
$i_1$ nor $i_j$ are dominated by the target of the conditional's else case.
In this example, new $\phi$-functions are not necessary: new variable
names are not alive together with the original variables.
The part (b) of this figure shows the solution that we get to this new
program.
The e-SSA form allows us to bind interval information directly to the live
ranges of variables; thus, giving us the opportunity to solve range analysis
sparsely.
More traditional approaches, which we call {\em dense analyses}, bind
interval information to pairs formed by variables and program points.

@@

\section{Related works}
\label{sec:rel}

% Early implementations of range analysis.
Range analysis is an old ally of compiler writers.
The notions of widening and narrowing were introduced by Cousot and Cousot in
one of the most cited papers in computer science~\cite{Cousot77}.
Different algorithms for range analysis have been later proposed
by Patterson~\cite{Patterson95}, Stephenson {\em et al.}~\cite{Stephenson00},
Mahlke {\em et al.}~\cite{Mahlke01} and many other researchers.
Recently there have been many independent efforts to find exact, polynomial
time algorithms to solve constraints on the interval
lattice~\cite{Costan05,Gawlitza09,Lakhdar11,Su04,Su05}.
However, these works are still very theoretical, and have not yet been used to
analyze large programs.

% Scalability: Astree and others.
There have been many practical approaches to abstract interpretation,
with special emphasis on range analysis~\cite{Bertrane10,Blanchet03,Cousot09,Gampe11,Jung05}.
Cousot's group, for instance, has been able to globally analyze programs with
thousands of lines of code, albeit they use domain specific tools.
Astr\'{e}e, for example, analyzes programs that do not contain recursive calls.
The work that is the closest to us is Oh {\em et al.}'s very recent abstract
interpretation framework~\cite{Oh12}.
Oh {\em et al.} discuss an implementation of range analysis on the interval
lattice that scales up to a program with 1,363K LoC (ghostscript-9.00), but
they do not provide results about its precision.
We could not find the benchmarks used in those experiments for a
direct comparison -- the distribution of ghostscript-9.00 available in the
LLVM test suite has 27K LoC.
On the other hand, we globally analyzed our largest benchmark, SPEC CPU 2006's
\texttt{gcc}, enabling function inlining, in less than 15 seconds.
Oh {\em et al.}'s implementation took orders of magnitude more time to go over
programs of similar size.
However, whereas they provide a framework to develop general sparse analyses, we
only solve range analysis on the interval lattice.

@@

In this presentation we chose to omit correctness proofs for our algorithms.
For a proof that the widening and the narrowing operators give origin to
approximating sequences, we recommend Cousot and Cousot's work~\cite{Cousot77}.
For a proof that the e-SSA form transformation is semantics preserving, we
invite the reader to check a recent report of ours~\cite{Tavares11b}.
In that paper we also show that the results that we obtain via the
sparse analysis are equivalent to the results provided by a dense analysis.
In other words, if the dense analysis tells us that variable $v$ is
associated with the interval range $[l, u]$ at program point $i$, and $v'$
is the new name of $v$, alive at $i$ in the e-SSA form program, then $v'$ is
associated with the interval $[l, u]$ along its entire live range.
Additionally, we have used a dynamic profiler, available in our webpage, to
empirically validate our results.

Our implementation of the range analysis algorithm described in this paper is
publicly available at \texttt{http://code.google.com/p/range-analysis/}.
This repository contains instructions about how to deploy and use our
implementation.
We provide a gallery of examples, including source codes,
control flow graphs and constraint graphs that we produce for meaningful
programs at \texttt{http://code.google.com/p/range-analysis/wiki/gallery}.

>>>>

We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
variable.
As we will show in Section~\ref{sub:micro}, once the growth pattern of $v$ is
known, we can replace $\fun{ft}(v)$ by an actual value.
After splitting the live ranges according to the rules stated above, we might
have to insert $\phi$-functions into the transformed program to re-convert it to
SSA form.
This last step avoids that two different names given to the same original
variable be simultaneously {\em alive} at the program code.
A variable $v$ is alive at a program point $p$ if the program's control flow
graph contains a path from $p$ to a site where $v$ is used, that does not
go across any re-definition of $v$.
Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
standard e-SSA form.
We have not created variable names for $i_f$ and $j_f$, because neither
$i_1$ nor $i_j$ are dominated by the target of the conditional's else case.
In this example, new $\phi$-functions are not necessary: new variable
names are not alive together with the original variables.
The part (b) of this figure shows the solution that we get to this new
program.
The e-SSA form allows us to bind interval information directly to the live
ranges of variables; thus, giving us the opportunity to solve range analysis
sparsely.
More traditional approaches, which we call {\em dense analyses}, bind
interval information to pairs formed by variables and program points.

@@

\section{Related work}
\label{sec:rel}

\noindent
\textbf{The Scalability of Range Analysis. }
% Early implementations of range analysis.
Range analysis is an old ally of compiler writers.
The notions of widening and narrowing were introduced by Cousot and Cousot in
one of the most cited papers in computer science~\cite{Cousot77}.
Different algorithms for range analysis have been later proposed
by Patterson~\cite{Patterson95}, Stephenson {\em et al.}~\cite{Stephenson00},
Mahlke {\em et al.}~\cite{Mahlke01} and many other researchers.
Recently there have been many independent efforts to find exact, polynomial
time algorithms to solve constraints on the interval
lattice~\cite{Gawlitza09,Su05,Costan05,Lakhdar11,Su04}.
However, these works are still very theoretical, and have not yet been used to
analyze large programs.

% Scalability: Astree and others.
There have been many practical approaches to abstract interpretation,
with special emphasis on range analysis~\cite{Gampe11,Blanchet03,Bertrane10,Cousot09,Jung05}.
Cousot's group, for instance, has been able to globally analyze programs with
thousands of lines of code, albeit using domain specific tools.
Astr\'{e}e, for example, analyzes programs that do not contain recursive calls.
The work that is the closest to us is Oh {\em et al.}'s very recent abstract
interpretation framework~\cite{Oh12}.
Oh {\em et al.} discuss an implementation of range analysis on the interval
lattice that scales up to a program with 1,363K LoC (ghostscript-9.00).
Because their focus is speed, they do not provide results about precision.
We could not find the benchmarks used in those experiments for a
direct comparison -- the distribution of ghostscript-9.00 available in the
LLVM test suite has 27K LoC.
On the other hand, we globally analyzed our largest benchmark, SPEC CPU 2006's
\texttt{gcc}, enabling function inlining, in less than 15 seconds.
Oh {\em et al.}'s implementation took orders of magnitude more time to go over
programs of similar size.
However, whereas they provide a framework to develop general sparse analyses, we
only solve range analysis on the interval lattice.

\noindent
\textbf{Sparse Data-Flow Analyses. }
In this paper we provide a {\em sparse} implementation of range analysis.
Sparsity, in our context, means that we associate points in the lattice of
interest -- intervals in our case -- directly to variables.
Dense analyses map such information to pairs formed by variables and program
points.
The compiler related literature contains many descriptions of sparse
data-flow analyses.
Some among these analyses obtain sparsity by using specific program
representations, like we did.
Others rely on data-structures.
In terms of data-structures, the first, and best known method proposed to
support sparse data-flow analyses is Choi {\em et al.}'s {\em Sparse Evaluation
Graph} (SEG)~\cite{Choi91}.
The nodes of this graph represent program regions where information produced by
the data-flow analysis might change.
Choi {\em et al.}'s ideas have been further expanded, for example, by Johnson
{\em et al.}'s {\em Quick Propagation Graphs}~\cite{Johnson93}, or Ramalingan's
{\em Compact Evaluation Graphs}~\cite{Ramalingan02}.
Building upon Choi's pioneering work, researchers have developed many
efficient ways to build such graphs~\cite{Pingali95,Pingali97,Johnson94}.
These data-structures have been shown to improve many data-flow analyses in
terms of runtime and memory consumption.
Nevertheless, the elegance of SEGs and its successors have not, so far, been
enough to attract the attention of mainstream compiler writers.
Compilers such as gcc, LLVM or Java Hotspot rely, instead, on several types of
program representations to provide support to sparse data-flow analyses.

Most eminent among these representations is the Static Single Assignment
form~\cite{Cytron91}, which suits well forward flow analyses, such as reaching
definitions.
Since its debut, the SSA form has been expanded in different ways.
For instance, the Gated SSA form allows the static association of logical
predicates with data-flow paths~\cite{Ottenstein90,Tu95}.
Scott Ananian has introduced in the late nineties the {\em Static Single
Information} (SSI) form, a program representation that supports both forward and
backward analyses~\cite{Ananian99}.
This representation was later revisited by Jeremy Singer~\cite{Singer06} and,
a few years later, by Boissinot {\em et al.}~\cite{Benoit09}.
Singer provided new algorithms plus examples of applications that benefit from
the SSI form, and Boissinot {\em et al.}, in an effort to clarify some
misconceptions about this program representation, introduced the notions of
{\em week} and {\em strong} SSA form.
Another important representation, which supports data-flow analyses that
acquire information from uses, is the \emph{Static Single Use} form (SSU).
There exists many variants of
SSU~\cite{Plevyak96,George03,Lo98}.
For instance, the ``strict'' SSU form enforces that each definition reaches a
single use, whereas SSI and other variations of SSU allow two consecutive uses
of a variable on the same path.
The program representation that we have used in this paper -- the {\em Extended
Static Single Assignment} (e-SSA) form -- was introduced by Bodik
{\em et al.}~\cite{Bodik00}.

There are so many different program representations because they fit specific
data-flow problems.
Each representation, given a domain of application, provides the following
property: the information associated with the live range of a variable is
invariant along every program point where this variable is alive.
There are two key aspects that distinguish one representation from the others:
firstly, where the information about a variable is acquired, and secondly,
how this information is propagated.
The e-SSA form, for instance, supports flow analyses that obtain information
both from variable definitions and conditional tests and propagate this
information forwardly.
Such analyses are also supported by the SSI form; hence, we could have used this
other representation too.
However, as we have shown in previous work, the e-SSA form is considerably
more economical~\cite{Tavares11b}.

@@

In this presentation we chose to omit correctness proofs for our algorithms.
For a proof that the widening and the narrowing operators give origin to
approximating sequences, we recommend Cousot and Cousot's work~\cite{Cousot77}.
For a proof that the e-SSA form transformation is semantics preserving, we
invite the reader to check a recent report of ours~\cite{Tavares11b}.
In that paper we also show that the results that we obtain via the
sparse analysis are equivalent to the results provided by a dense analysis.
In other words, if the dense analysis tells us that variable $v$ is
associated with the interval $[l, u]$ at program point $i$, and $v'$
is the new name of $v$, alive at $i$ in the e-SSA form program, then $v'$ is
associated with the interval $[l, u]$ along its entire live range.
Additionally, we have used a dynamic profiler, available in our webpage, to
empirically validate our results.

Our implementation of the range analysis algorithm described in this paper is
publicly available at \texttt{http://code.google.com/p/range-analysis/}.
This repository contains instructions to deploy and use our
implementation.
We provide a gallery of examples, including source codes,
control flow graphs and constraint graphs that we produce for meaningful
programs at \\ \texttt{http://code.google.com/p/range-analysis/wiki/gallery}.
